Case Study: RNA-Seq Differential Analysis (basal ganglia)

### 2018-10-30 datasets/GSEA/fgsea-human.Rmd
Added data description and replaced dropbox links for zenodo links.

The objective of this document is to compare different methods for multiple testing correction in the context of Gene Set Enrichment Analysis (GSEA). In order to get a list of genes, we will use RNA-seq data consisting of cortex and cerebellum samples of a subset of the GTEx data.

Here we use the `fgsea` Bioconductor package to implement the GSEA method. This is 
a Functional Class Scoring approach, which does not require setting an arbitrary threshold
for Differential Expression, but instead relies on the gene's rank (here we rank by
DESeq2 test statistic).

@@

The dataset consists of 10 samples (6 cerebellum and 4 cortex samples), all from male individuals. We download a *DESeqDataSet* if not present locally and modify the design to test for differences in gene expression between cerebellums and cortex.

```{r, message=FALSE}
if (!file.exists(count_file)) {
    download.file("https://www.dropbox.com/s/itrs2a6cnmurnck/gsea-human-counts.rds?dl=1",
                  destfile = count_file)
}
dsdObject <- readRDS(count_file)
design( dsdObject ) <- ~ tissue
```

>>>>

The objective of this document is to compare different methods for multiple testing correction in the context of Gene Set Enrichment Analysis (GSEA). We use RNA-seq data consisting of cortex and cerebellum samples of a subset of the GTEx individuals. Shortly, samples were downloaded using the *Short Read Archive Toolkit* and mapped to the human reference genome version GRCh38 using *STAR v2.4.2a*. *htseq-count* was used to tabulate the number of uniquely mapping reads for each gene. We used the `DESeq2` package to format the data into a `DESeqDataSet` object.

We use the `fgsea` Bioconductor package to implement the GSEA method. This is 
a Functional Class Scoring approach, which does not require setting an arbitrary threshold for Differential Expression, but instead relies on the gene's rank (here we rank by DESeq2 test statistic).

@@

We download the *DESeqDataSet* from zenodo, if not present locally, and modify the design to test for differences in gene expression between cerebellums and cortex.

```{r, message=FALSE}
if (!file.exists(count_file)) {
    download.file("https://zenodo.org/record/1475409/files/gsea-human-counts.rds?download=1",
                  destfile = count_file)
}
dsdObject <- readRDS(count_file)
design( dsdObject ) <- ~ tissue
```

### 2018-10-30 datasets/GSEA/fgsea-mouse.Rmd
Added data description and replaced dropbox links for zenodo links.

The objective of this vignette is to test different multiple testing methods in the context of Gene Set Enrichment Analysis (GSEA). To do this, we will use mouse data from the paper by [Cabezas-Wallscheid et al. (Cell stem Cell, 2014)](https://www.sciencedirect.com/science/article/pii/S1934590914003014?via%3Dihub). The dataset consists of RNA-seq data from mouse hematopoietic stem cells and 4 multipotent progenitor lineages. We will test for difference in gene expression between two different cell fractions and perform GSEA in the resulting list of genes.

@@

```{r, message=FALSE}
if (!file.exists(count_file)) {
    download.file("https://www.dropbox.com/s/jq6giqeclxcdypq/gsea-mouse-counts.rds?dl=1",
                  destfile = count_file)
}
dseHSCMPP <- readRDS(count_file)
```

@@

```{r gocat, message=FALSE}
if (!file.exists(goset_file)) {
    mart <- useMart("ensembl", "mmusculus_gene_ensembl")
    goSets <- getBM(c("ensembl_gene_id", "go_id"), mart = mart,
                    filters = "ensembl_gene_id", values = rownames(res))
    goSets <- goSets[!nchar(goSets$go_id) == 0, ]
    goSets <- with(goSets, split(go_id, ensembl_gene_id))
    saveRDS(goSets, file = goset_file)
} else {
   goSets <- readRDS(goset_file)
}
```

>>>>

The objective of this vignette is to test different multiple testing methods in the context of Gene Set Enrichment Analysis (GSEA). To do this, we use data from the paper by [Cabezas-Wallscheid et al. (Cell stem Cell, 2014)](https://www.sciencedirect.com/science/article/pii/S1934590914003014?via%3Dihub). The data consist of RNA-seq data from mouse hematopoietic stem cells and multipotent progenitor lineages. The raw fastq data is available through the ArrayExpress database (http://www.ebi.ac.uk/arrayexpress) under accession number E-MTAB-2262. These data were mapped to the mouse reference genome GRCm38 (ENSEMBL release 69) using the Genomic Short-Read Nucleotide Alignment program (version 2012-07-20). We used htseq-count to count the number of reads overlapping with each gene and used the DESeq2 package to format the data as a DESeqDataSet R object.

@@

```{r, message=FALSE}
if (!file.exists(count_file)) {
    download.file("https://zenodo.org/record/1475409/files/gsea-mouse-counts.rds?download=1",
                  destfile = count_file)
}
dseHSCMPP <- readRDS(count_file)
```

@@

```{r gocat, message=FALSE}
if (!file.exists(goset_file)) {
    library(biomaRt)
    mart <- useMart("ensembl", "mmusculus_gene_ensembl")
    goSets <- getBM(c("ensembl_gene_id", "go_id"), mart = mart,
                    filters = "ensembl_gene_id", values = rownames(res))
    goSets <- goSets[!nchar(goSets$go_id) == 0, ]
    goSets <- with(goSets, split(go_id, ensembl_gene_id))
    saveRDS(goSets, file = goset_file)
} else {
   goSets <- readRDS(goset_file)
}
```

### 2018-10-30 datasets/RNA-Seq/RNAseq-gtex.Rmd
Added data description and replaced dropbox links for zenodo links.

The first dataset consists of 10 samples each from two regions of the human basal ganglia, the nucleus accumbens and the putamen, obtained from the GTEx project. Shortly, samples were downloaded using the \emph{Short Read Archive Toolkit} and mapped to the human reference genome version GRCh38 using \emph{STAR v2.4.2a}. \emph{htseq-count} was used to tabulate the number of uniquely mapping reads for each gene.

@@

We download a DESeqDataSet object containing gene level counts for GTEx samples from the nucleus accumbens and the putamen.

```{r, dsd-download}
if (!file.exists(count_file)) {
    download.file("https://www.dropbox.com/s/9g06cx9vyonhdao/rnaseq-brain-counts.rds?dl=1",
                  destfile = count_file)
}
dsd <- readRDS(count_file)
```

>>>>

The data consists of 20 samples from two regions of the human basal ganglia, the nucleus accumbens and the putamen, from the GTEx project. Shortly, samples were downloaded using the *Short Read Archive Toolkit* and mapped to the human reference genome version GRCh38 using *STAR v2.4.2a*. *htseq-count* was used to tabulate the number of uniquely mapping reads for each gene. We used *DESeq2* to format the data into a *DESeqDataSet* object.

@@

We download the DESeqDataSet object from zenodo that contains the gene level counts for GTEx samples.

```{r, dsd-download}
if (!file.exists(count_file)) {
    download.file("https://zenodo.org/record/1475409/files/rnaseq-brain-counts.rds?download=1", destfile = count_file)
}
dsd <- readRDS(count_file)
```
